Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services

### 2024-10-12 manuscript/01-introduction.qmd
Introduction revised according to Prof Bahman comments

Universal access  essential medicines is a cornerstone of effective healthcare systems, yet ensuring equitable availability of these vital medications remains an urgent global challenge[@quick2003essential;@world2004annual]. Across continents, especially in Africa and Asia, many nations grapple with restricted access due to factors like unaffordability and suboptimal supply chain management [@world2004medicines]. A survey conducted across eight sub-Saharan African countries indicated that there were  unacceptably low availability of essential medicines.The mean availability of 12 priority essential medicines for women ranged from 22% to 40% and from 28% to 57% for children [@droti2019poor]. However, the issue of drug shortages extends beyond developing countries, with instances reported in the United States and Europe, adversely impacting healthcare systems and compromising patient care quality [@fox2003managing, @kaakeh2011impact; @johnson2011drug; @huys2013european;@le2011prevalence].

The consequences of drug shortages are felt both financially within the healthcare sector and in terms of the quality of patient care [@alspach2012drug; @kaakeh2011impact;@kaakeh2011impact; @baumer2004national].The consequences extend across patients, providers, healthcare institutions, and research programs. Patients face delays or cancellations of treatments, substandard care due to the unavailability of optimal therapies, and increased costs in a secondary "grey market." Compromised patient safety, complications, prolonged hospital stays, and even deaths have been reported [@alspach2012drug;].Addressing these shortages is crucial to preserving the quality of patient care and ensuring the effectiveness of healthcare delivery. Thus, a smoothly operating pharmaceutical supply chain is essential to ensure timely access to essential medicines, making accurate demand forecasting an indispensable component.

An efficient pharmaceutical supply chain is vital for timely access to essential health commodities. One of the most crucial components of such a supply chain is a well-defined demand forecasting process. Forecasting accurately the future demand for essential medicines provides valuable insights for anticipating future healthcare requirements including procurement and replenishment of inventory[@subramanian2021effective]. Despite the significance of forecasting for supply chain management, pharmaceutical product forecasting remains a complex endeavor influenced by a multitude of factors. Key considerations include the availability of data, the specific position within the supply chain, and the nuanced understanding of contributing elements. The position within the supply chain introduces unique challenges and dynamics, requiring tailored forecasting approaches. Additionally, a nuanced understanding of contributing elements involves grasping intricate factors that impact forecasting accuracy, such as market trends, regulatory changes, and external influences. Schneider, Wilson, and Rosenbeck (2010) and Hyndman and Athanasopoulos (2018) highlight the intricate nature of pharmaceutical forecasting, emphasizing the need for a comprehensive approach that incorporates these multifaceted aspects to enhance supply chain management effectiveness[@schneider2010pharmaceutical; @hyndman2018forecasting].

The Ethiopian pharmaceutical sector's distinct characteristics, marked by a diverse array of medicines and unique data challenges stemming from service delivery points, staff capacity limitations, communication hurdles, prolonged procurement and distribution lead times, poor forecast accuracy, and policy intricacies, necessitates tailored forecasting strategies that cater to the nation's specific context [@boche2022procurement]. In this context, a comprehensive exploration of effective forecasting models that align with Ethiopia's pharmaceutical landscape is paramount to mitigating stock outs and enhancing healthcare delivery. The implications of pharmaceutical demand forecasting extend beyond financial considerations to societal well-being [@rostami2022forecasting]. In Ethiopia, a country undergoing demographic and healthcare evolution, precise demand forecasting assumes immense importance for resource allocation, supply chain optimization, and policy formulation. With mounting interest in refining demand forecasting in Ethiopia's pharmaceutical sector, this study seeks to contribute to existing knowledge by investigating and recommending proficient demand forecasting methods. Our investigation centers on the application of mean, naive, exponential smoothing, regression, and ARIMA models to bolster forecast accuracy and effectiveness. This paper aims to forecast the demand for selected pharmaceuticals in EPSS over the next 12 months by utilizing various forecasting models and integrating predictors. The integration of predictors with the models will be examined to determine whether they enhance the forecasting accuracy. These predictors are derived from a historical analysis of events in Ethiopia over the last 5 years that have impacted pharmaceutical sales in EPSS. These events include COVID-19, conflicts in the country, disease campaigns, budget closures, budget releases, and physical inventory, all of which, in varying degrees, influence the demand for pharmaceuticals in Ethiopia.

The current paper offers the following contributions we aim at forecasting pharmaceutical sales over a 12-month period these models aim to enhance the accuracy and reliability of demand predictions within the context of EPSS. Moreover an essential contribution lies in the identification of key factors influencing pharmaceutical sales. The study develops a model that effectively incorporates these driving factors, providing a more nuanced and comprehensive approach to demand prediction. We generate forecasts and evaluate model performance using the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error (RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error (MASE), Crips and Winkler are used to assess probabilistic forecast accuracy. 
#how you know the forecasting accutacy problem and provide eveidence about the forecasting problems 
# most forecast focus on point forecast and we want to focus on probablistic forecast
# i will write the contribution of the reearch rather than these objectives

This research is guided by several objectives:

1. Evaluate the robustness of various forecasting models for pharmaceutical demand prediction within EPSS.
2. Identify pharmaceuticals that exhibit superior forecast accuracy under specific models.
3. Assess the performance of forecast accuracy metrics to gauge the efficacy of selected models.
4. Test the incorporation of diverse predictors affecting forecasts and evaluate resulting forecasting accuracy.
In pursuing these objectives, this study endeavors to elevate the precision and reliability of pharmaceutical demand forecasting, thereby contributing to optimized supply chain management and enhanced healthcare outcomes in Ethiopia.

>>>>

Universal access to essential medicines is a fundamental goal of effective healthcare systems, yet ensuring equitable availability continues to be a critical global challenge [@quick2003essential; @world2004annual]. Many countries, particularly in Africa and Asia, face significant barriers to accessing these medications, often due to high costs and inefficient supply chain management [@world2004medicines]. A survey in eight sub-Saharan African countries revealed concerningly low availability of essential medicines, with mean availability ranging from 22% to 40% for women’s priority medicines and from 28% to 57% for children’s medicines [@droti2019poor]. However, drug shortages are not limited to low-income regions; they also occur in developed areas like the United States and Europe, where they disrupt healthcare delivery and compromise the quality of patient care [@fox2003managing; @kaakeh2011impact; @johnson2011drug; @huys2013european; @le2011prevalence].

The consequences of drug shortages extend beyond immediate healthcare delivery, impacting healthcare costs, treatment quality, and patient safety [@alspach2012drug; @kaakeh2011impact; @baumer2004national]. Such shortages can result in treatment delays, suboptimal care due to the unavailability of preferred therapies, increased reliance on costly secondary markets, and serious patient outcomes, including complications, longer hospital stays, and even deaths [@alspach2012drug]. Addressing these challenges requires a resilient pharmaceutical supply chain that can ensure the timely availability of medicines. In this context, accurate demand forecasting becomes critical for effective procurement and inventory management [@subramanian2021effective].

Demand forecasting in pharmaceutical supply chains is a complex endeavor, influenced by factors such as data quality, supply chain positioning, and external drivers, including market trends and regulatory changes [@schneider2010pharmaceutical; @hyndman2018forecasting]. While vast amounts of data are often collected for transactional purposes—such as sales—many forecasting approaches overlook the significance of domain knowledge, which is as crucial as quantitative data for understanding the context and making more informed predictions. Domain expertise helps identify the impact of nuanced factors, such as clinical guidelines, disease outbreaks, and local policies, that are not always evident from sales data alone.

Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview]. Domain knowledge enables analysts to incorporate relevant factors, such as market conditions, policy changes, and supply chain disruptions, into the forecasting model. It aids in guiding feature selection, parameter tuning, and the interpretation of complex data patterns, ultimately leading to more accurate and actionable forecasts [@sinha2008incorporating; @dash2022review]. This integration is particularly beneficial in fields like healthcare, where expert insights can align machine learning algorithms with real-world applications and ethical considerations [@dash2022review].

Despite the considerable body of research on integrating domain knowledge with statistical techniques, relatively few studies focus specifically on the unique requirements of pharmaceutical forecasting. This study aims to bridge that gap by employing advanced machine learning models combined with domain expertise to address the distinctive characteristics of Ethiopia’s pharmaceutical sector. The challenges in this sector, such as diverse product categories, limited data from service delivery points, communication issues, long lead times, forecast inaccuracies, and complex policies, necessitate tailored forecasting approaches [@bilal2024challenges; @boche2022procurement]. Accurate demand forecasting in Ethiopia is essential for optimizing resource allocation, streamlining supply chain operations, and shaping effective healthcare policies [@rostami2022forecasting].

This research will forecast the demand for selected pharmaceuticals in the Ethiopian Pharmaceutical Supply Service (EPSS) over the next six months using various methods, including dynamic regression ARIMA, multiple regression, neural networks, machine learning, and time-series GPT models. We will explore whether incorporating domain knowledge can improve forecast accuracy by integrating predictors derived from historical events over the past five years, such as stock replenishment cycles, stockouts, fiscal calendar variations, COVID-19, conflicts, and seasonal trends like malaria outbreaks.

Our contributions include not only generating point forecasts but also employing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods. Additionally, we identify key factors influencing pharmaceutical demand and incorporate these into our forecasting models to better reflect the complexities of the Ethiopian context. Model performance will be evaluated using metrics such as Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.

The remainder of this paper is structured as follows: Section 2 reviews the literature and identifies gaps to position our work; Section 3 outlines the experimental design, including data sources, forecasting methods, and evaluation criteria; Section 4 presents the results and analysis; and Section 5 concludes with a summary of findings and future research directions.
